A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-07-31 papers/water-saft/paper.tex
Updated water-saft stuff for new paper Tested everything on node09.

\fixme{Some stuff about water}.  An important aspects of water is its
ability to form hydrogen bonds.  Each water molecule contains two
hydrogen bond donor sites and two acceptor sites.

\subsection{Statistical Associating Fluid Theory}

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation} \label{eq:SAFT-free-energy}
  F = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms---ideal gas, hard-sphere repulsion and
dispersion---encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy for fluids that are chain polymers. In our model, we use a hard
sphere reference fluid which renders $F_\textit{chain}$ irrelivant for
our functional.

In a recent paper \cite{hughes2013classical}, Hughes \emph{et al}
constructed a classical density functional theory for water which used
Clark \emph{et al}'s \cite{clark2006developing} five fitting
paramenters for an optimal cDFT for water.  In addition, they
introduced a parameter for the length scale over which the dispersion
interaction is correlated

@@

Schulte \emph{et al}\cite{schulte2012using} derived expressions for
the correlation fucntoin at contact based on the White Bear version of
FMT. They found their $g_{\sigma}^A$ to agree with Monte Carlo
simulation than Yu and Wu's which we was used in Hughes \emph{et al}'s
paper.  We use Hughs \emph{et al}'s functional and replace the
correlation with Schulte \emph{et al}'s, which we will call
$g_{\sigma}$. \fixme{expand this}.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/surface-tension}
\end{center}
\caption{Comparison of Surface tension versus temperature for theoretical and
  experimental data. The experimental data is taken from NIST.\cite{nistwater}
  The length-scaling parameter $s_d$ is fit so that the theoretical surface 
  tension will match the experimental surface tension near room temperature.}
\label{fig:surface-tension}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonding per molecule at
  distances from the center of a hydrophobic rod for various
  radii. The solid lines are the results from using the older
  $g_{\sigma}$, while the dotted lines are from Schulte \emph{et al}'s
  asymmetric $g_{\sigma}$.}
\label{fig:single-rod-X}
\end{figure}

>>>>

Water \fixme{stuff about water}.  An important aspects of water
is its ability to form hydrogen bonds.  Each water molecule contains
two hydrogen bond donor sites and two acceptor sites.

One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature.
Classical DFT is a natural framework for creating a more flexible
theory of hydrophobicity that can readily describe interaction of
water with arbitrary external potentials---such as potentials
describing strong interactions with solutes or surfaces.

Statistical Associating Fluid Theory (SAFT) is a theory describing
complex fluids in which hydrogen bonding plays a significant
role\cite{chapman1989saft, muller2001molecular}. SAFT is used to
accurately model the equations of state of both pure fluids and
mixtures over a wide range of temperatures and pressures. In a recent
paper \cite{hughes2013classical}, we constructed a classical DFT for
water which uses SAFT as a basis for a free energy functional. We
used Clark \emph{et al}'s \cite{clark2006developing} five fitting
parameters for an optimal cDFT for water and, in addition, introduced
a parameter for the length scale over which the dispersion interaction
is correlated.

SAFT is based on Wertheim's first-order thermodynamic perturbation
theory (TPT1)\cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV}, which allows it to
account for strong associative interactions between
molecules. Wertheim's thermodynamic perturbation theory uses a
hard-sphere reference fluid where each sphere has a number of
association sites on the surface. The association sites have an
effective range and a small energy contribution when a site is within
range of another hard-sphere's site. The free energy due to this
association depends on the radial distribution function
$g_\sigma^\text{HS}(\rr)$. Our previous water functional used Yu and
Wu's radial distribution function
\cite{yu2002fmt-dft-inhomogeneous-associating}, which was found to
agree less with hard-sphere Monte-Carlo simulation than a distribution
function developed by us in a recent paper.\cite{schulte2012using} In
our paper, we developed an approximation for correlation at
contact to describe the hard-sphere fluid.

@@

We modify our previous the association free energy term to use our
$g_{\sigma}^A$ rather than Yu and Wu's $g_\sigma^{YuWu}$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{previous paper}.
  \end{cases} 
\end{align}
The four in $n_{site}(\rr)$ comes from the four hydrogen bond sites
and, $n_o(\rr)$ is the average density at contact, and $\zeta(\xx)$ is
a dimensionless measure of the density inhomogeneity from Yu and
Wu. \cite{yu2002fmt-dft-inhomogeneous-associating} The functional
$X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

We use Clark's five empirical parameters, and fit the calculated
surface tension to experimental surface tension at ambient conditions.
We find these agree when $\lscale$ is 0.454.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/surface-tension}
\end{center}
\caption{Comparison of Surface tension versus temperature for theoretical and
  experimental data. The experimental data is taken from NIST.\cite{nistwater}
  The length-scaling parameter $\lscale$ is fit so that the theoretical surface 
  tension will match the experimental surface tension near room temperature.}
\label{fig:surface-tension}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-compare}
\end{center}
\caption{ Density profiles for single rods of different
  radii for the new functional (top) and the old functional
  (bottom). Each separately colored line represents a density profile
  for a rod; the black dotted line is the bulk density.}
\label{fig:density-single-rod}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of hydrophobic rods of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}^{HS}$.}
\label{fig:single-rod-X}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-broken-HB}
\end{center}
\caption{ Broken hydrogen bonds broken when rods are placed in
  homogeneous water.  \fixme{Our old and new functional are both plotted here}}
\label{fig:single-rod-broken-HB}
\end{figure}

\section{Hydrogen bonding near a hydrophobic rod}

We first simulate a single hydrophobic rod immersed in water. The
density profile of the water with our functional is compared with that
of Hughes in Fig. \ref{fig:density-single-rod}. The density at contact
with the rods are nearly identical, even more so for larger radius
rods. Since we have changed the association term in the functional, a
difference will occur in the average number of hydrogen bonds in a
given system. We compare the hydrogen bonds with the new and old
functional for rods of a few different radii
(Fig. \ref{fig:single-rod-X}) and find that there is a notable
difference.

\section{Hydrogen bonding near hard-sphere solutes}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/spheres-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of spherical cavities of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}$.}
\label{fig:spheres-X}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/sphere-broken-HB}
\end{center}
\caption{ \fixme{spheres!}}
\label{fig:spheres-broken-HB}
\end{figure}

We continue our study into three dimensions with a hard sphere solute
in water.\fixme{stuff}.

We have modified a classical DFT for water with a more accurate radial
distribution function which changes the way that the molecules assosciate.
